Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-09-20 paper/main.tex
add more details to model description

We use principles from discrete differential geometry~\cite{bobenko2008discrete} to describe the
stiffness and flexibility of the bacterial rod.
As explained previously, MReB will counteract the bending of the rod, resulting in a stiffening
force.
In principle, this stress will be dependent on the position inside the cell~\cite{Chatterjee1988}
but in order to construct a model which can be parametrized and by method of Occam's Razor, we chose
to model the bacterium as a bending beam, thus utilizing the simplest solution which can still
produce the desired effects.
Such a beam can be represented by a collection of vertices $\vec{x}_i$ which are connected by a
springs of lengths $l_i$.
We use the notation $\vec{c}_i:=\vec{x}_i-\vec{x}_{i-1}$ with which the resulting force acting
between two vertices if given by

@@

To simulate proliferation, we introduce a growth term for the spring lengths $l_i$.
In most of our simulations, we choose to use identical lengths for all segments (i.e. $l_i=l$).
However, in order to generalize our approach we assume that individual lengths of segments could
vary.
We reduce the growth rate with increasing number of neighbors $N$.
This behaviour can be toggled on or off thus removing the additional term in brackets of
equation~\ref{eq:growth-rod-lengths}.

@@

Note that the growth rate is proportional to the rod length $l$ which leads to exponential growth.
This process will increase the length of the cell indefenitely unless we impose a condition for the
division event.
We define a fixed length threshold for the total length of the cell at which it divides.
To construct a new cell, we cannot simply copy the existing one twice, but we also need to adjust
internal parameters in the process.
The following actions need to be taken for the old and new agent.
We first calculate the new spring lengths of the two resulting rods.
\begin{equation}
    \tilde{l} = l_i\left(\frac{1}{2} - \frac{r}{\sum\limits_i l_i}\right)
    \label{eq:growth-rod-new-lengths}
\end{equation}
Afterwards, we start at one endpoint $\vec{x}_0$ and travel along the polygon defined by the
bacterial rod.
We stop when we have reached a length that matches the length of the new spring length $\tilde{l}$.
From here, we continue and iterate this procedure until we have found all new vertices $\vec{w}_i$.
Equation~\ref{eq:growth-rod-new-lengths} ensures that that the tips of the two new rods are touching
right after the division event.

@@

We chose to optimize for the damping of agents' movement within the medium, 3 interaction parameters
of the Mie potential $n,m,V_0$ a single radius for all agents and growth rates which were chosen
individually for each cell.
We removed any stochastic motion contributions by setting $D=0$.
This leads to a total of $5+1\times 6=11$ parameters.
To estimate these parameters we used the differential evolution~\cite{Storn1997} global optimization
algorithm which was initially seeded with a sample generated by the latin-hypercube~\cite{McKay1979}
method.
Finally, we performed a local minimization using the L-BFGS-B algorithm~\cite{Liu1989}.

>>>>

We use principles from discrete differential geometry~\cite{bobenko2008discrete} to describe the
stiffness and flexibility of the bacterial rod~\cite{Amir2014_2}.
As explained previously, MReB will counteract the bending of the rod, resulting in a stiffening
force.
In principle, this stress will be dependent on the position inside the cell~\cite{Chatterjee1988}
but in order to construct a model which can be parametrized and by method of Occam's Razor, we chose
to model the bacterium as a bending beam, thus utilizing the simplest solution which can still
produce the desired effects.
Such a beam can be represented by a collection of vertices $\vec{x}_i$ which are connected by a
springs of lengths $l_i$.
We use the notation $\vec{c}_i:=\vec{x}_i-\vec{x}_{i-1}$ with which the resulting force acting
between two vertices if given by

@@

To simulate proliferation, we introduce a growth term for the spring lengths $l_i$.
In most of our simulations, we choose to use identical lengths for all segments (i.e. $l_i=l$).
However, in order to generalize our approach we assume that individual lengths of segments could
vary.
Our implemented model allows us to reduce the growth rate with increasing number of neighbors
$n\leq N$.
This option provides a phenomenological approach for modeling slower growth in larger colonies but
is be toggled off by default thus removing the additional term in brackets of
equation~\ref{eq:growth-rod-lengths}.

@@

Note that the growth rate is proportional to the rod length $l$ which leads to exponential
growth~\cite{Robert2014,Takeuchi2005}.
This process will increase the length of the cell indefenitely unless we impose a condition for the
division event.
We define a fixed length threshold for the total length of the cell at which it divides.
To construct a new cell, we cannot simply copy the existing one twice, but we also need to adjust
internal parameters in the process.
The following actions need to be taken for the old and new agent.
We first calculate the new spring lengths of the two resulting rods.
\begin{equation}
    \tilde{l} = l_i\left(\frac{1}{2} - \frac{r}{\sum\limits_i l_i}\right)
    \label{eq:growth-rod-new-lengths}
\end{equation}
Afterwards, we start at one endpoint $\vec{x}_0$ and travel along the polygon defined by the
bacterial rod.
We stop when we have reached a length that matches the length of the new spring length $\tilde{l}$.
From here, we continue and iterate this procedure until we have found all new vertices $\vec{w}_i$.
Equation~\ref{eq:growth-rod-new-lengths} ensures that that the tips of the two new rods are touching
right after the division event.

@@

We chose to optimize for the damping of agents' movement within the medium, 3 interaction parameters
of the Mie potential $n,m,V_0$ a single radius for all agents and growth rates which were chosen
individually for each cell.
We removed any stochastic motion contributions by setting $D=0$.
This leads to a total of $5+1\times 6=11$ variable parameters.
In order to model non-flexible rods, we set the rigidity and spring tension of the rod to a high,
fixed value.
To estimate these parameters we used the differential evolution~\cite{Storn1997} global optimization
algorithm which was initially seeded with a sample generated by the latin-hypercube~\cite{McKay1979}
method.
Finally, we performed a local minimization using the L-BFGS-B algorithm~\cite{Liu1989}.
